Can we distinguish modes of selective interactions using linkage disequilibrium?

### 2021-03-23 Manuscript.Rmd
consolidate figures, incorporate comments

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium between selected variants to test for selective interference and
epistatic interactions, with some disagreement over the interpretation of
observations from data. Interpreting this data is hindered by the relative lack
of analytic or even numerical expectations for patterns of variation between
pairs of loci under the combined effects of selection, dominance, epistasis,
and demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations.
Finally, I use whole-genome sequencing data from humans to assess how well we
can differentiate modes of selective interactions in practice. I find that
positive LD between missense mutations within genes is driven by strong
positive correlations between pairs of mutations that fall within the same
conserved domain, pointing to compensatory mutations or antagonistic epistasis
as the prevailing mode of interaction within but not outside of conserved genic
elements. The heterogeneous landscape of both mutational fitness effects and
selective interactions within protein-coding genes calls for more refined
inferences of the joint distribution of fitness and interactive effects, and
the methods presented here should prove useful in that pursuit.

@@

In their foundational paper, @Hill1966-gv recognized that linked selected
variants reciprocally impede the efficacy of selection at each locus. Linked
selection reduces the fixation probability of advantageous mutations and
increases that of deleterious mutations compared to expectations under
single-locus models [@Birky1988-jm]. Allele frequency dynamics and correlations
of linked selected variants are also predicted to deviate from expectations
without interference. Under a multiplicative fitness model, where the fitness
reduction of a haplotype carrying multiple deleterious variants is equal to the
product of the fitness reduction of each mutation independently, we should
expect to see those mutation segregate on different haplotypes more often than
together, leading to negative, or repulsion, linkage disequilibrium (LD),
although the extent of LD depends non-trivially on the strength of selection
and probability of recombination separating loci [@Hill1966-gv;
@McVean2000-ox].

@@

@Sohail2017-zq observed negative LD between loss-of-function variants in
protein-coding genes (such as stop gains and losses, frameshifts, and other
nonsense mutations) in both human and fruit fly populations, which they
interpreted as evidence for widespread synergistic epistasis between these
mutations. Within the past year or two, both @Sandler2020-vk and @Garcia2020-zn
have reevaluated patterns of LD between coding variants in humans, fruit flies,
and *Capsella grandiflora*, and suggested interference and dominance may
instead be driving patterns of LD [@Garcia2020-zn] or questioned whether LD
between loss-of-function variants is significantly different from zero
[@Sandler2020-vk].

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution
stores the expected density or observed counts of pairs of biallelic loci with
each possible haplotype frequency configuration in a sample of size $n$. In
many cases it is simpler to visualize summaries (such as $\E[D]$) or slices of
the distribution (Figure \@ref(fig:fig1)) instead of the full three-dimensional
distribution. Here, I focus on low-order LD statistics, such as $\E[D]$ and
$\E[D^2]$, and their decay with recombination distance as these are statistics
that are commonly used to test for interactions between loci. Instead of
unnormalized $\E[D^2]$ and $\E[D]$, I consider expectations for $\sigma_d^2 =
\E[D^2]/\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$
and $q$ are the derived allele frequencies at the left ($A/a$) and right
($B/b$) loci, respectively. These normalized statistics have two benefits:
first, the mutation rate cancels so that expectations are robust to assumptions
about the per-base mutation rate, and second, analytic expectations for these
quantities are available under neutrality at constant population size for
comparison [@Ohta1971-yd].

@@

@@

For mutations under additive selection (i.e. $h=1/2$) and no epistasis, we
recover the well known Hill-Robertson interference result of negative LD
between selected mutations, which is strongest for pairs for mutations that
have selection coefficients $\gamma=O(1)$, or $s \approx 1/2N_e$ (Figure
\@ref(fig:fig1)D). For strongly deleterious mutations, LD is close to zero even
with tight linkage, as they almost always segregate at low enough frequencies
that they are unlikely to interfere with each other [@McVean2000-ox].

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig2)). Synergistic epistasis results in negative LD while
antagonistic epistasis results in positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \@ref(fig:fig2)D even moderately deleterious mutations with
population-size-scaled selection coefficients of $\gamma=-10$ show large mean
LD that extends well beyond 100 kilobases. More strongly deleterious
interacting mutations are expected to show large signed LD over much larger
recombination distances.

### Dominance and interference

@@

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial, and some parameter regimes can cause strong negative LD
between pairs of negatively selected variants. For strong selection, pairs of
recessive mutations ($h<1/2$) can have large negative signed LD (Figure
\@ref(fig:fig3)D), although it decays rapidly with increasing distance between
loci and is roughly zero for $\rho \gg 1$. For weakly selected mutations with
$\gamma=-1$, we see that there is no monotonic effect of the level of dominance
on negative LD, with both recessive and dominant pairs of mutations having more
negative LD than the additive case.

@@

While dominance and epistasis are well-characterized models of selection, a
large family of selection models can be specified by the assigning unique
selection coefficients to possible diploid pairs of haplotypes. If we assume
the diploid genotype that is homozygous for the ancestral allele ($ab/ab$) has
fitness 1, then there are nine other possible diploid two-locus genotypes that
could be given unique fitnesses (Table \@ref(tab:selmodels)), noting that
$AB/ab$ and $Ab/aB$ genotypes can have differing fitness effects.

The case with $s_{AB/ab}\not=s_{Ab/aB}$ can arise in a scenario where a
mutation at either locus within a haplotype impacts some functional region or
element, but a diploid individual carrying at least one copy that is free of
mutations has minimal fitness loss. This is analogous to the single-locus
dominance case, where $s_{Aa} < \frac{1}{2}s_{AA}$. In this "gene-based
dominance" scenario [e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher
fitness than an $Ab/aB$ type (a simple implementation of this model is given in
the final column in Table \@ref(tab:selmodels)). Such a gene-based fitness
model gives similar expected positive signed LD to the model of antagonistic
epistasis (compare Figure \@ref(fig:fig4)E-F to G-H), although the
interpretation of those two models can differ. It shows that with a highly
parameterized space of possible general diploid selection models, multiple
models with different biological interpretations can give similar patterns of
expected $D$.

@@

The moment system for $\Psi_n$ readily incorporates variable population size as
well as mutation and recombination rates and selection parameters that change
over time. Here I focus on non-equilibrium population size history and consider
scenarios relevant to human demographic history, including bottlenecks and
expansions. I explore two simple models (Figure \@ref(fig:fig4)A), one with an
instantaneous expansion and another with a bottleneck followed by recovery. I
also consider two demographic histories inferred using genome-wide genealogy
reconstruction [@Speidel2019-nj] applied to the
@1000_Genomes_Project_Consortium2015-zq dataset, and focus on size histories
for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
ancestry (CEU) (Figure \@ref(fig:fig4)B).

For each of the four histories, Figure \@ref(fig:fig4)C-H shows the dynamics of
$\sigma_d^1$ for a given parameterization of two-locus selection, including
synergistic and antagonistic epistasis, dominance within loci, and gene-based
dominance. In general across each selection model, population size expansions
are not expected to strongly affect $\sigma_d^1$, whether that expansion occurs
deeper in the past as in the simple expansion model or rapid expansion more
recently, as in the YRI. On the other hand, population size reductions push
signed LD to more extreme values and subsequent recoveries or expansion again
reduce the magnitude of LD. Under no selection condition tested here do
population size changes cause expected LD to change sign, showing that while
the magnitude of deviation of LD from zero is sensitive to population size
history, interpreting general patterns of the observed sign of LD in data
should not be strongly affected by population size history.

@@

Synonymous and missense mutations show similar levels of positive LD when
considering pairs of mutations within the same gene. Loss-of-function mutations
instead have more negative LD, suggesting differing modes of selective
interactions for loss-of-function and missense mutations (Figure
\@ref(fig:fig5)). Within each population, large measurement noise gives 95\%
confidence intervals that overlap with zero across mutation class, although the
pattern is remarkably consistent across African, European, and East Asian
populations in the Thousand Genomes dataset. Comparing mean LD within genes in
African and Eurasian populations, LD in Eurasian populations is somewhat larger
on average, that is, more positive for missense mutations and more negative for
loss-of-function mutations. This is in agreement with differences in
expectations between populations that have or have not gone through a
bottleneck in their recent past (Figure \@ref(fig:fig4)). For pairs of missense
mutations in different protein-coding genes, signed LD decays to zero fairly
rapidly and is lower than signed LD within genes even for mutations that are
linked (distance less that 50 kb).

@@

To further explore the observation of positive LD between both synonymous and
missense mutations within genes, I examine the role that annotated conserved
domains in protein-coding genes play in driving this signal. Note that the low
density of loss-of-function mutations within annotated domains in humans
precludes analysis of LD between those mutations at the level of sub-elements
within genes. Protein-coding domains are conserved elements of genes, often
associated with some known functional or structural feature of a protein. As
such, we would expect that purifying selection is stronger on average within
domains than within the same gene but outside of a conserved domains. Indeed,
the site-frequency spectrum (SFS) is skewed to lower frequencies for both
missense and loss-of-function mutations within domains when compared to the
same classes of mutations outside of domains, with much more negative values of
Tajima's D within domains (Table \@ref(tab:tajimasD)). On the other hand, no
difference is observed for synonymous mutations whether within or outside
domains, suggesting roughly equivalent effects of selection (either direct or
linked) on synonymous variation.

For pairs of missense mutations that both fall within the same functional
domain, we observe large positive LD, roughly an order of magnitude larger than
pairs of synonymous mutations that both fall within the same domain (Figure
\@ref(fig:fig6)). This difference between pairs of missense and synonymous
variants within domains is observed for both tightly linked ($< 10$bps) and
more distant ($>1$kb) pairs (Figure \@ref(fig:lddecaydomain)). Assuming most
synonymous mutations are neutral, we would expect signed LD between missense
mutations to be less than that between synonymous mutations under predictions
from models of either Hill-Robertson interference, synergistic epistasis, or
dominance (Figures \@ref(fig:fig1)-\@ref(fig:fig3)). This observation is
opposite to those expectations, suggesting a different prevailing interactive
effect between nonsynonymous mutations within domains than those that are
commonly tested or modeled.

The large positive signal of LD for missense mutations within the same
domain does not extend to pairs of missense mutations that are in
different domains. Pairs of missense and synonymous mutations show
nearly equal levels of LD close to zero across domains, with missense mutations
even slightly more negative than synonymous mutations (Figure \@ref(fig:fig6)).
The interactive effect driving large LD within domains is therefore likely
domain-specific. However, the average distance between pairs of mutations
within domains is much smaller than between domains, so this observation may be
primarily driven by the higher recombination distances between pairs of mutations
across distinct domains.

@@

I also measure LD between pairs of mutations that both fall outside of
annotated domains. For such mutations, I consider averages across all pairs of
mutations outside of domains, as well as subsets of pairs matched to the same
distances as pairs of mutation within and between domains as surveyed in the
previous section (Figures \@ref(fig:fig6)B-D and \@ref(fig:domainspops)).
Averaging over all pairs of mutations within the same gene but outside domains,
both synonymous and missense mutations have LD slightly above zero, with
synonymous mutations observed to have slightly higher signed LD than missense
mutations. This pattern is primarily driven by tightly linked variants. For
pairs of mutations at distances matched to those within the domains, synonymous
mutations have much larger positive LD than missense mutations, a signal that
vanishes for pairs matched to distances between domains. For larger distances,
both synonymous and missense mutations have LD approximately zero.

@@

We do not see an increase in LD between mutations that are found in different
annotated domains within the same gene, which are on average at much larger
recombination distances. Nor do we find increased LD outside of domains.
Rather, for pairs of mutations outside of domains but at the same distance as
those within the same domain, missense variants have considerably lower LD than
than synonymous variants. This difference between synonymous and missense pairs
of variants largely disappears for SNPs separated by more than a few hundred or
a thousand base pairs (Figure \@ref(fig:lddecay)), which suggests that
Hill-Robertson interference is the primary mode of interaction between
mutations falling outside of domains, as epistasis is expected to impact LD
over larger distances than what is observed.

@@

@@

@@

Instead, closed, numerical approaches to compute expectations for two-locus
statistics under a wide range of selection models, as presented here or in
@Ragsdale2019-nt and @Friedlander2021-bs, allow us to explore this highly
parameterized space of models far more efficiently, and it opens the
possibility for performing likelihood-based inference using signed LD or other
two-locus summaries of the data. For example, inferring the joint distribution
of dominance and selection is underpowered using the SFS alone, but because
signed LD is sensitive to the levels of dominance (Figure \@ref(fig:fig3)),
inferring the DFE with dominance may be feasible using the joint distribution
of allele frequencies and LD.

@@

To obtain the recursion equation under selection we consider drawing $n$
lineages from generation $t$, which has has an expected sampling distribution
of haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
derived haplotype, each of those sampled lineages has probability of being
rejected equal to the relative fitness of its haplotype state. If a lineage is
rejected, a replacement is drawn from the full population. Under the assumption
that $ns_{\cdot\cdot} \ll 1$, the probability that more than one selection
event occurs in any given generation is negligibly small, so that the case of
multiple simultaneous rejections can be ignored. Then $Psi_n^{t+1}$ relies only
on $\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full selection
operator $\mathcal{S}$ for additive selection is given in the Supporting
Information.

@@

# Acknowledgements
- Discussions: Simon Gravel, Alex Diaz-Papkovich, Matthias Steinrücken, Eric Friedlander
- Comments on ms:

Asked from comments from
- Kevin Thornton
- Mashaal Sohail
- Simon Gravel
- Matthias Steinrücken

\clearpage
\newpage

# Figures

```{r fig1, echo=F, fig.cap="(ref:fig1cap)", fig.align="center"}
knitr::include_graphics("figures/fig1.pdf")
```

>>>>

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium between selected variants to test for selective interference and
epistatic interactions, with some disagreement over the interpretation of
observations from data. Interpreting this data is hindered by the relative lack
of analytic or even numerical expectations for patterns of variation between
pairs of loci under the combined effects of selection, dominance, epistasis,
and demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations.
Finally, I use whole-genome sequencing data from humans to assess how well we
can differentiate modes of selective interactions in practice. I find that
positive LD between missense mutations within genes is driven by strong
positive allele-frequency correlations between pairs of mutations that fall
within the same conserved domain, pointing to compensatory mutations or
antagonistic epistasis as the prevailing mode of interaction within but not
outside of conserved genic elements. The heterogeneous landscape of both
mutational fitness effects and selective interactions within protein-coding
genes calls for more refined inferences of the joint distribution of fitness
and interactive effects, and the methods presented here should prove useful in
that pursuit.

@@

In their foundational paper, @Hill1966-gv recognized that linked selected
variants reciprocally impede the efficacy of selection at each locus, a process
known as selective interference. Linked selection reduces the fixation
probability of advantageous mutations and increases that of deleterious
mutations compared to expectations under single-locus models [@Birky1988-jm].
Allele frequency dynamics and correlations of linked selected variants are also
predicted to deviate from expectations without interference. Under a
multiplicative fitness model, where the fitness reduction of a genome carrying
multiple deleterious variants is equal to the product of the fitness reduction
of each mutation independently, we expect to see net linkage disequilibrium
(LD) equal to zero for unlinked sites [@Kondrashov1995-va]. But for linked
loci, those mutations are expected to segregate on different haplotypes more
often than together, leading to negative, or repulsion, LD, although the extent
of LD depends non-trivially on the strength of selection and probability of
recombination separating loci [@Hill1966-gv; @McVean2000-ox].

@@

@Sohail2017-zq observed negative LD between loss-of-function variants in
protein-coding genes (such as stop gains and losses, frameshifts, and other
nonsense mutations) in both human and fruit fly populations. They interpreted
this as evidence for widespread synergistic epistasis between these mutations,
in which the fitness reduction of multiple mutations is greater than the
product of that of each individual mutations independently. Within the past
year or two, both @Sandler2020-vk and @Garcia2020-zn have reevaluated patterns
of LD between coding variants in humans, fruit flies, and *Capsella
grandiflora*, and suggested interference and dominance may instead be driving
patterns of LD [@Garcia2020-zn] or questioned whether LD between
loss-of-function variants is significantly different from zero
[@Sandler2020-vk].

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution
stores the expected density or observed counts of pairs of biallelic loci with
each possible haplotype frequency configuration in a sample of size $n$. In
many cases it is simpler to visualize summaries (such as $\E[D]$) or slices of
the distribution (Figures \@ref(fig:fig1) and \@ref(fig:hudson) ) instead of
the full three-dimensional distribution. Here, I focus on low-order LD
statistics, such as $\E[D]$ and $\E[D^2]$, and their decay with recombination
distance as these are statistics that are commonly used to test for
interactions between loci. Instead of unnormalized $\E[D^2]$ and $\E[D]$, I
consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$ and $q$ are the derived allele
frequencies at the left ($A/a$) and right ($B/b$) loci, respectively. These
normalized statistics have two benefits: first, the mutation rate cancels so
that expectations are robust to assumptions about the per-base mutation rate,
and second, analytic expectations for these quantities are available under
neutrality at constant population size for comparison [@Ohta1971-yd].

@@

```{r fig1, echo=F, fig.cap="(ref:fig1cap)", fig.align="center", fig.pos="tbp!"}
knitr::include_graphics("figures/fig1.pdf")
```

(ref:fig1cap) \textbf{Additive selection and epistasis.} (A and B) The
distribution at stationarity of $AB$ haplotype counts in a sample size of 30,
in which we observe 10 $A$ alleles at the left locus, and 10 $B$ alleles at the
right locus. (C) The decay of normalized squared LD ($\sigma_d^2$) with
recombination distance ($\rho=4N_er$) for pairs of variants with a given
selection coefficient. The dashed line is the neutral expectation from
@Ohta1971-yd, given in Equation \@ref(eq:ohta). (D) The decay of
$\sigma_d^1=\E[D]/ \E[p(1-p)q(1-q)]$ for pairs of variants. The negative LD
effect of Hill-Robertson interference is strongest for intermediate selection,
with $\gamma\sim-1$ to $-5$, where $\gamma=4N_es$. **[The text does not refer
to specific panels, 1c should give the equation for $\sigma_d^2$, and not sure
what we get from 1a and 1b. Define rho in caption, and place other parameters
in parentheses]** **[Move a, b to supp, consolidate c and d with fig 2]** And
from figure 2 previously ...(A, B) Synergistic epistasis decreases
$\sigma_d^2$, while antagonistic epistasis increases $\sigma_d^2$, though both
are decreased in comparison to neutral expectations. (C, D) Epistasis expected
signed LD away from zero, an effect which can extend over large recombination
distances. Here, $\gamma$ is the selection coefficient for derived alleles $A$
and $B$, assuming selection is equal at the two loci. $\epsilon=0$ is the case
of pure additive selection with no epistasis, and $\gamma_{AB}=(\gamma_A +
\gamma_B)(1+\epsilon)$. Dashed lines are expectations under neutrality for both
$\sigma_d^2$ and $\sigma_d^1$. **[More explicitly define that positive e is
synergistic, negative is antagonistic, and define formula in the methods]**
**[caption for C, D reads a little trailing]**

@@

For mutations under additive selection (i.e. $h=1/2$) and no epistasis, we
recover the well known Hill-Robertson interference result of negative LD
between selected mutations, which is strongest for pairs of mutations that have
selection coefficients $\gamma = O(1)$, or $s \approx 1/2N_e$ (Figure
\@ref(fig:fig1)B). For strongly deleterious mutations, LD is close to zero even
with tight linkage, as they almost always segregate at low enough frequencies
that they are unlikely to interfere with each other [@McVean2000-ox].

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig1)). Synergistic epistasis results in negative LD while
antagonistic epistasis results in positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak **[example in supp]**.
Epistasis-induced LD can extend over long distances, especially for strongly
deleterious mutations. For example, in Figure \@ref(fig:fig1)F even moderately
deleterious mutations with population-size-scaled selection coefficients of
$\gamma=-10$ show large mean LD that extends well beyond 100 kilobases. More
strongly deleterious interacting mutations are expected to show large signed LD
over much larger recombination distances.

### Dominance

@@

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial, and some parameter regimes can cause strong negative LD
between pairs of negatively selected variants. For strong selection, pairs of
recessive mutations ($h<1/2$) can have large negative signed LD (Figure
\@ref(fig:fig2)D), although it decays rapidly with increasing distance between
loci and is roughly zero for $\rho \gg 1$. For weakly selected mutations with
$\gamma=-1$, we see that there is no monotonic effect of the level of dominance
on negative LD, with both recessive and dominant pairs of mutations having more
negative LD than the additive case.

```{r fig2, echo=F, fig.cap="(ref:fig2cap)", fig.align="center", fig.pos="tbp!"}
knitr::include_graphics("figures/fig2.pdf")
```

(ref:fig2cap) **Dominance effects.** (A-B) Pairs of recessive mutations
($h<0.5$) have higher squared LD than additive mutations, while pairs of
dominant alleles have lower expected $D^2$. (C-D) Pairs of recessive, strongly
deleterious mutations can have large negative LD, although this correlation
decays more rapidly with larger recombination than LD under epistasis (Figure
\@ref(fig:fig1)D and F). Notably, there is no simple relationship between strength of
selection and dominance, particularly when $s$ is $O(1/N)$. **[Epistasis is
zero in this setup, and rearrange panels to match figure 1]**

@@

**[Struggled a bit with this section, make clearer and easier to parse]**

While dominance and epistasis are well-characterized models of selection, a
large family of selection models can be specified by assigning unique
selection coefficients to possible diploid pairs of haplotypes. If we assume
the diploid genotype that is homozygous for the ancestral allele ($ab/ab$) has
fitness 1, then there are nine other possible diploid two-locus genotypes that
could be given unique fitnesses (Table \@ref(tab:selmodels)), noting that
$AB/ab$ and $Ab/aB$ genotypes can have differing fitness effects.

The case with $s_{AB/ab}\not=s_{Ab/aB}$ can arise in a scenario where a
mutation at either locus within a haplotype impacts some functional region or
element, but a diploid individual carrying at least one copy that is free of
mutations has minimal fitness loss. In this "gene-based dominance" scenario
[e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher fitness than an $Ab/aB$
type (a simple implementation of this model is given in the final column in
Table \@ref(tab:selmodels)). Such a gene-based fitness model gives similar
expected positive signed LD to the model of antagonistic epistasis (compare
Figure \@ref(fig:fig3)E-F to G-H), although the interpretation of those two
models can differ. It shows that with a highly parameterized space of possible
general diploid selection models, multiple models with different biological
interpretations can give similar patterns of expected $D$.

@@

The moment system for $\Psi_n$ readily incorporates variable population size as
well as mutation and recombination rates and selection parameters that change
over time. Here I focus on non-equilibrium population size history and consider
scenarios relevant to human demographic history, including bottlenecks and
expansions. I explore two simple models (Figure \@ref(fig:fig3)A), one with an
instantaneous expansion and another with a bottleneck followed by recovery. I
also consider two demographic histories inferred using genome-wide genealogy
reconstruction [@Speidel2019-nj] applied to the
@1000_Genomes_Project_Consortium2015-zq dataset, and focus on size histories
for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
ancestry (CEU) (Figure \@ref(fig:fig3)B).

For each of the four histories, Figure \@ref(fig:fig3)C-H shows the dynamics of
$\sigma_d^1$ for a given parameterization of two-locus selection, including
synergistic and antagonistic epistasis, dominance within loci, and gene-based
dominance. In general across each selection model, population size expansions
are not expected to strongly affect $\sigma_d^1$, whether that expansion occurs
deeper in the past as in the simple expansion model or rapid expansion more
recently, as in the YRI. On the other hand, population size reductions push
signed LD to more extreme values and subsequent recoveries or expansion again
reduce the magnitude of LD. Under no selection condition tested here do
population size changes cause expected LD to change sign, showing that while
the magnitude of deviation of LD from zero is sensitive to population size
history, interpreting general patterns of the observed sign of LD in data
should not be strongly affected by population size history.

```{r fig3, echo=F, fig.cap="(ref:fig3cap)", fig.align="center", fig.pos="tbp!"}
knitr::include_graphics("figures/fig3.pdf")
```

(ref:fig3cap) **The effects of demography on signed LD.** Left column:
simulations under two toy models, one with an instantaneous expansion in the
past, and the other with a 10-fold reduction in size followed by recovery.
Right column: simulations under population size histories inferred using Relate
[@Speidel2019-nj] applied to populations from
@1000_Genomes_Project_Consortium2015-zq. In all simulations, $\gamma$ is set to
$-2$ at both loci. (A-B) Model illustrations. Note the differences in
scale. (C-D) Additive selection with no epistasis, showing fully linked
($\rho=0$) and more loosely linked ($\rho=5$) pairs. Bottleneck events result
in larger negative values of signed LD, but population size changes are not
expected to change the sign of mean LD. (E-F) Synergistic ($\epsilon > 0$) and
antagonistic ($\epsilon < 0$) epistasis. Again, bottleneck events result in
larger signed LD. Here, $\epsilon=1/2$ is a model with $\gamma_A=-2$,
$\gamma_B=-2$ and $\gamma_{AB}=-6$, and $\epsilon=-1/2$ is a model with
$\gamma_A=-2$, $\gamma_B=-2$ and $\gamma_{AB}=-2$. (G-H) Dominance within loci
results in negative LD, while dominance acting at the gene level results in
positive LD. Here, $h=0.1$. Again, bottlenecks result in larger average LD in
both cases.

@@

Synonymous and missense mutations show similar levels of positive LD when
considering pairs of mutations within the same gene. Loss-of-function mutations
instead have more negative LD, suggesting differing modes of selective
interactions for loss-of-function and missense mutations (Figure
\@ref(fig:fig4)). Within each population, large measurement noise gives 95\%
confidence intervals that overlap with zero across mutation class, although the
pattern is remarkably consistent across African, European, and East Asian
populations in the Thousand Genomes dataset. Comparing mean LD within genes in
African and Eurasian populations, LD in Eurasian populations is somewhat larger
on average, that is, more positive for missense mutations and more negative for
loss-of-function mutations. This is in agreement with differences in
expectations between populations that have or have not gone through a
bottleneck in their recent past (Figure \@ref(fig:fig3)). For pairs of missense
mutations in different protein-coding genes, signed LD decays to zero fairly
rapidly and is lower than signed LD within genes even for mutations that are
linked (distance less that 50 kb).

```{r fig4, echo=F, fig.cap="(ref:fig4cap)", fig.align="center", fig.pos="tbp!"}
knitr::include_graphics("figures/fig4.pdf")
```

(ref:fig4cap) **LD within and between protein-coding genes.** (A-B) Pairs of
variants in coding regions partitioned by classes of mutations.
Loss-of-function mutations are defined as stop and start gains and losses and
other nonsense mutations of high impact. In all populations, we observe
positive mean signed LD for missense and synonymous variants, although the
confidence interval of each measurement overlaps with zero. Similarly, we
observe mean negative LD for pairs of loss-of-function variants within genes.
There are only $\sim200-300$ pairs of loss-of-function variants that fall
within the same gene across the populations tested here, causing large
uncertainty in each measurement. Gray lines indicate 2 standard errors of the
measurement, computed by bootstrapping of sets of genes. (C-F) While pairs of
synonymous and missense mutations within genes are consistently positive, pairs
between genes are not consistent in their sign across populations. LD for all
populations analyzed here is shown in Figure \@ref(fig:genespops).

@@

To further explore the observation of positive LD between both synonymous and
missense mutations within genes, I examine the role that annotated conserved
domains in protein-coding genes play in driving this signal. Protein-coding
domains are conserved elements of genes, often associated with some known
functional or structural feature of a protein. As such, we would expect that
purifying selection is stronger on average within domains than within the same
gene but outside of a conserved domains. Indeed, the site-frequency spectrum
(SFS) is skewed to lower frequencies for both missense and loss-of-function
mutations within domains when compared to the same classes of mutations outside
of domains, with much more negative values of Tajima's D within domains (Table
\@ref(tab:tajimasD)). On the other hand, no difference is observed for
synonymous mutations whether within or outside domains, suggesting roughly
equivalent effects of selection (either direct or linked) on synonymous
variation.

```{r fig5, echo=F, fig.cap="(ref:fig5cap)", fig.align="center", fig.pos="tbp!"}
knitr::include_graphics("figures/fig5.pdf")
```

(ref:fig5cap) **LD within and between annotated domains in protein-coding
genes.** (A) Considering only variants that fall within annotated domains in
protein-coding genes, pairs of mutations that fall within the same gene display
larger positive LD than pairs that lie in different domains. This pattern is
most striking for missense mutations, for which pairs within domains have
average signed LD that is consistently larger than for pairs that span two
domains, or for pairs of synonymous mutations that are within the same domain.
(B-D) The slightly positive LD that we observe for missense and synonymous
mutations gene-wide (Figure \@ref(fig:fig3)A) is largely driven by the subset
of pairs mutations that fall within the same annotated domain. LD for all
populations analyzed here isshown in Figure \@ref(fig:domainspops).

For pairs of missense mutations that both fall within the same functional
domain, we observe large positive LD, roughly an order of magnitude larger than
pairs of synonymous mutations that both fall within the same domain (Figure
\@ref(fig:fig5)). This difference between pairs of missense and synonymous
variants within domains is observed for both tightly linked ($< 10$bps) and
more distant ($>1$kb) pairs (Figure \@ref(fig:lddecaydomain)). Assuming most
synonymous mutations are neutral, we would expect signed LD between missense
mutations to be less than that between synonymous mutations under predictions
from models of either Hill-Robertson interference, synergistic epistasis, or
dominance (Figures \@ref(fig:fig1) and \@ref(fig:fig2)). This observation is
opposite to those expectations, suggesting a different prevailing interactive
effect between nonsynonymous mutations within domains than those that are
commonly tested or modeled.

The large positive signal of LD for missense mutations within the same
domain does not extend to pairs of missense mutations that are in
different domains. Pairs of missense and synonymous mutations show
nearly equal levels of LD close to zero across domains, with missense mutations
even slightly more negative than synonymous mutations (Figure \@ref(fig:fig5)).
The interactive effect driving large LD within domains is therefore likely
domain-specific. However, the average distance between pairs of mutations
within domains is much smaller than between domains, so this observation may be
primarily driven by the higher recombination distances between pairs of mutations
across distinct domains.

@@

I also measure LD between pairs of mutations that both fall outside of
annotated domains. For such mutations, I consider averages across all pairs of
mutations outside of domains, as well as subsets of pairs matched to the same
distances as pairs of mutation within and between domains as surveyed in the
previous section (Figures \@ref(fig:fig5)B-D and \@ref(fig:domainspops)).
Averaging over all pairs of mutations within the same gene but outside domains,
both synonymous and missense mutations have LD slightly above zero, with
synonymous mutations observed to have slightly higher signed LD than missense
mutations. This pattern is primarily driven by tightly linked variants. For
pairs of mutations at distances matched to those within the domains, synonymous
mutations have much larger positive LD than missense mutations, a signal that
vanishes for pairs matched to distances between domains. For larger distances,
both synonymous and missense mutations have LD approximately zero.

@@

We do not see an increase in LD between mutations that are found in different
annotated domains within the same gene, which are on average at much larger
recombination distances. Nor do we find increased LD outside of domains.
Rather, for pairs of mutations outside of domains but at the same distance as
those within the same domain, missense variants have considerably lower LD than
synonymous variants. This difference between synonymous and missense pairs of
variants largely disappears for SNPs separated by more than a few hundred or a
thousand base pairs (Figure \@ref(fig:lddecay)). This suggests that
Hill-Robertson interference is the primary mode of interaction between
mutations falling outside of domains, as epistasis is expected to impact LD
over larger distances than what is observed, although the strength of epistasis
is also likely to be a function of distance between mutations.

@@

**[Mashaal points out that simulations with structure leading to inbreeding, in which subpopulations have differing load, can lead to positive LD]**

@@

**[Need to point out that confidence intervals are wide, and while the signal
is consistent across populations, shared evolutionary history makes formal
testing of significance difficult.]**

@@

Instead, closed, numerical approaches to compute expectations for two-locus
statistics under a wide range of selection models, as presented here or in
@Ragsdale2019-nt and @Friedlander2021-bs, allow us to explore this highly
parameterized space of models far more efficiently, and it opens the
possibility for performing likelihood-based inference using signed LD or other
two-locus summaries of the data. For example, inferring the joint distribution
of dominance and selection is underpowered using the SFS alone, but because
signed LD is sensitive to the levels of dominance (Figure \@ref(fig:fig2)),
inferring the DFE with dominance may be feasible using the joint distribution
of allele frequencies and LD.

@@

To obtain the recursion equation under selection we consider drawing $n$
lineages from generation $t$, which has an expected sampling distribution
of haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
derived haplotype, each of those sampled lineages has probability of being
rejected equal to the relative fitness of its haplotype state. If a lineage is
rejected, a replacement is drawn from the full population. Under the assumption
that $ns_{\cdot\cdot} \ll 1$, the probability that more than one selection
event occurs in any given generation is negligibly small, so that the case of
multiple simultaneous rejections can be ignored. Then $Psi_n^{t+1}$ relies only
on $\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full selection
operator $\mathcal{S}$ for additive selection is given in the Supporting
Information.

@@

# Acknowledgements

I am grateful to Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal
Sohail, Matthias Steinrücken, and Kevin Thornton for helpful discussions and
feedback on earlier versions of this manuscript that improved this work.
